Verification of concurrent design patterns with data

### 2019-02-22 2018-Arch-pNets/arch-pNets.tex
presque pret...
arch-pNets paper: Clean-up before intermediate submission to Coordination '19

\paragraph{Encoding the coordinators.}
The encoding of a coordinator $C$ is a pLTS with: 1)~an
 initial state and an   
$\init$ transition that initialises all the variables to those defined
by the initial valuation $\val{0}{C}$ of $C$, 2)~an action algebra that matches the actions of the coordinator ports but adds, an additional boolean action parameter, and also, for each exported
variable $x$, a corresponding fresh input variable $?x'$ to allow updates during interactions, 3)~pLTS transitions that reflect the original transitions of
$C$ with $\true$ as parameter and 4)~additional loop transitions marked by
$\false$.
\noteLH{enleve le 3 et integre au 2}
Formally,  $\nopri[C] \bydef{=} \pNetTuple{S, s_0, \goesto[{\nopri[C]}]{}}$, such that
%
\begin{itemize}
\item $s_0 \not\in Q_C$ is fresh and $S = Q_C \cup \{s_0\}$,
  \noteEM{En principe les $vars(s)$ devraient etre disjoints, mais je
    veux bien laisser ca sous le tapis tant qu'on n'implemente pas le codage...}
\item $\vars(s) = V_C \cup \setdef{?x'}{x \in \export[C](p), p \in a, s\goesto{a} }$, for all $s \in Q_C$, and $\vars(s_0) =
  \emptyset$,
%% \item the transition relation $\goesto[{\nopri[C]}]{}$ is defined as
%%   the union of three sets of transitions
%%   (with $\mathit{init} \bydef{=} \bigl(x := \val{0}{C}(x)\bigr)^{x \in V_C}$):
%% %
%%   \begin{multline}
%%     \label{eq:nopri:trans}
%%     {\goesto[{\nopri[C]}]{}} \bydef{=}
%%     \bsetdef{\tuple{s, a(\true), g, u, s'}}{
%%       \tuple{s, a, g, u, s'} \in \goesto{}}
%%     \\    
%%     \cup
%%     \bsetdef{\tuple{s, a(\false), \true, \noop, s}}{
%%       s \in Q_C, a \in P_C}
%%     %% \\
%%     \cup
%%     \bigl\{(s_0, \init, \true, \mathit{init}, q^0_C)\bigr\}
%%     \,.
%%   \end{multline}

\item let  $\mathit{uinit} \bydef{=} \bigl(x := \val{0}{C}(x)\bigr)^{x \in V_C}$ and,
  \noteSB{Ludo, Eric, pourriez-vous v\'erifier que cela est
    compr\'ehensible ? Ok pour moi (Eric) OK Ludo, rephrase un peu}%
  \add{for all $s \goesto{a, g, u} s'$ with $u = (x\! :=
\! e_x)^{x
    \in V}$ (with $V \!\subseteq\! V_C$),
%  
  let $u' \bydef{=}
  \bigl(x := e_x\bigr)^{x \in V \setminus \export[C](a)}
  \cup
  \bigl(x := e_x\bigl[\nicefrac{?x'}{x}\bigr]\bigr)^{x \in V \cap \export[C](a)}$,
%
  and $\export[C]'(a) \bydef{=} \setdef{?x'}{x \in \export[C](p), p \in a}$,
%
\noteEM{j'ai inverse les $\export[C]'(a), \export[C](a)$ dans les
  actions, pour etre conforme a la notation de la Def 1 page 4. }
\begin{multline*}
    {\goesto[{\nopri[C]}]{}} \bydef{=} 
    \Setdef{\Tuple{s, a\bigl(\export[C]'(a), \export[C](a),\true\bigr), g, u', s'}}{
      s \goesto{a, g, u} s'}
    \\
    {} \cup
    \Setdef{\Tuple{s, a\bigl(\export[C]'(a), \export[C](a),\false\bigr), \true, \noop, s}}{
      s \in Q_C, \exists s' \in Q_C: s' \goesto{a}}
    \\
    {} \cup \bigl\{(s_0, \init, \true, \mathit{uinit}, q^0_C)\bigr\}.
  \end{multline*}}
\end{itemize}\vspace{-3ex}

@@

Basic research on behaviour models and verification algorithms for
data-sensitive systems started in the nineties, with the seminal work
of Hennesy, Lin, and their colleagues on value-passing systems with
assignments
\cite{CONCUR::Lin1996,HennessyL1995:353,APSEC::Lin2001}. Later, many
different works addressing various classes of infinite-state systems
and/or paramaterized topologies have been published, using
combinations of approaches, often including predicate abstraction and
SMT satisfiability (see
e.g. \cite{BruniEtAl-Tiles-Concur2000,DBLP:conf/cade/GhilardiNRZ08,DBLP:journals/jsat/AlbertiGPRR12,CimattiEtAl-NUXMV-CAV2014,ChampionEtAl-Kind2-CAV2016}). With
respect to these, we use symbolic representations not only to get a
finite representation of infinite spaces, but also to express the
(data-sensitive) synchronisations with the environment, making our
models suitable for compositional verification.
Among these works, several have shown the capacity of the SMT engines
(either Z3 or Yikes) as servers for solving verification conditions of
the algorithms, for large case-studies
(e.g. \cite{DBLP:journals/corr/CimattiGMT13,DBLP:journals/corr/abs-1806-11459}). Also 
%% SMT on an application not using component structure. is there a strong Z3 example == Eric

@@

  Denoting \todoSB{One tilde is enough}$\val[\doubletilde]{}{1}$ the restriction of
%
  $\val[\tilde]{}{}$ to
  $V_1 = V_{C_1} \cup \bigcup_{B \in \cB} V_B$,
%
  we have, by \eq{substitution} and the monotonicity of
  expressions,
%
  $(\val[\primeit]{}{1}, \val[\primeit]{}{}) \order
  \val[\doubletilde]{}{1}[e^1; (e_1, e_b)]$.  \todoSB{Remove this}\hl{Notice that all
  variables affected by both $e^1$ and $e^2$ must necessarily
  belong to $\bigcup_{B \in \cB} V_B$.}  Hence,
%
  $\val[\doubletilde]{}{1}[e^1; (e_1, e_b)] =
  (\val[\primeit]{}{1}, \val[\secondit]{}{})$ with 
  $\val[\primeit]{}{} \order \val[\secondit]{}{}$.

@@

  In particular, the first safety property \eq{prop:reset:can}:
  \begin{equation*}
    \AG \bigl(\PortFail(true) \rightarrow \EX \EU[]{\lnot \PortFail(true)}{\PortReset}\bigr).
  \end{equation*}
is not true, because the $OT_2$ loop on state $\ostate{11}$ stays in
state 11 where you need another $\PortFail(true)$ before enabling a
reset. This is because without priorities we can have an
activity on the dandling ``fail'' port without synchronising with the
others, even if they are available.

@@

\subsubsection{Scaling-up: composing architectures.}

Here is the code of the corresponding composed pNet:
\noteEMin{Il faudrait (mais a-t-on le temps?): faire un dessin de la
  composition au niveau BIP, et au niveau pNet. Un resultat brut, vaguement
  correct, montre 37 OTs, modulo les petites questions de mon mail de
  mercredi midi...}

\todoSBin{Ecrire les proprietes qu'il doit verifier... ? }

\bigskip
\begin{verbatim}
SV0: <tick(z, t, z1), fail(zone1, b11), fail(zone2, b12), fail1(b01), fail2(b02), _> 
     --> fail(b01, b02), 
     [(z1 = z |Intersection| ((b11?zone1:TopInt)+t) |Intersection| ((b12?zone2:TopInt)+t))
      /\(b11=>b01)/\(b12=>b02)]
SV1: <cancel(b1), resume(b21), resume(b22), resume1, resume2, _>-->resume, 
     [(b1=b21)/\(b1=b22)]
SV2: <timeout, timeout, timeout, _, _, ask>-->ask
SV3: <_, reset, reset, _, _, reset>-->reset
SV4: <init, init, init, _, _, _>-->init

pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
- Transition: t1---cancel(false)-->t1, {}
- Transition: t1---cancel(true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}

pLTS: FC1
- Transition: s0---init-->s1, { zone1 := minZone1,maxZone1}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false)-->s1, {}
- Transition: s1---fail(zone1, false)-->s1, {}
- Transition: s1---fail(zone1, true)-->s2, {}
- Transition: s2---fail(zone1, false)-->s2, {}
- Transition: s2---resume(false)-->s2, {}
- Transition: s2---resume(true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone1, false)-->s3, {}
- Transition: s3---timeout-->s3, {}
- Transition: s3---resume(false)-->s3, {}
- Transition: s3---reset-->s1, {}

pLTS: FC2
- Transition: s0---init-->s1, { zone2 := minZone2,maxZone2}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false)-->s1, {}
- Transition: s1---fail(zone2, false)-->s1, {}
- Transition: s1---fail(zone2, true)-->s2, {}
- Transition: s2---fail(zone2, false)-->s2, {}
- Transition: s2---resume(false)-->s2, {}
- Transition: s2---resume(true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone2, false)-->s3, {}
- Transition: s3---timeout-->s3, {}
- Transition: s3---resume(false)-->s3, {}
- Transition: s3---reset-->s1, {}

INFO  BIPArchFailureTimerMaxV4X2 - Total: 37 SATISFIABLE OTs
INFO  BIPArchFailureTimerMaxV4X2 -        3136 UNSATISFIABLE OTs
INFO  BIPArchFailureTimerMaxV4X2 - Total Time: 9111 ms
\end{verbatim}

\bigskip
\Fig{ArchFailure:OA-TimerV4X2} shows the open automata of
this composed architecture.
Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
false. As in the previous case, this is because we have not use here
the Max Progress hypothesis, otherwise all the ``*'' in the figure
should be replaced by ``$ff$''.

  \begin{figure}[t]
  \centering
  \includegraphics[width=0.8\columnwidth]{TimerV4X2}
  \caption{The open Automaton of the composed system}
  \label{fig:ArchFailure:OA-TimerV4X2}
\end{figure}

>>>>

\paragraph{Encoding the coordinators.}
The encoding of a coordinator $C$ is a pLTS with: 1)~an
 initial state and an   
$\init$ transition that initialises all the variables to those defined
by the initial valuation $\val{0}{C}$ of $C$, 2)~an action algebra that matches the actions of the coordinator ports but adds, an additional boolean action parameter, and also, for each exported
variable $x$, a corresponding fresh input variable $?x'$ to allow updates during interactions, 3)~pLTS transitions that reflect the original transitions of
$C$ with $\true$ as parameter and 4)~additional loop transitions marked by
$\false$.
%% \noteLH{enleve le 3 et integre au 2}
Formally,  $\nopri[C] \bydef{=} \pNetTuple{S, s_0, \goesto[{\nopri[C]}]{}}$, such that
%
\begin{itemize}
\item $s_0 \not\in Q_C$ is fresh and $S = Q_C \cup \{s_0\}$,
  %% \noteEM{En principe les $vars(s)$ devraient etre disjoints, mais je
  %%   veux bien laisser ca sous le tapis tant qu'on n'implemente pas le codage...}
\item $\vars(s) = V_C \cup \setdef{?x'}{x \in \export[C](p), p \in a, s\goesto{a} }$, for all $s \in Q_C$, and $\vars(s_0) =
  \emptyset$,
%% \item the transition relation $\goesto[{\nopri[C]}]{}$ is defined as
%%   the union of three sets of transitions
%%   (with $\mathit{init} \bydef{=} \bigl(x := \val{0}{C}(x)\bigr)^{x \in V_C}$):
%% %
%%   \begin{multline}
%%     \label{eq:nopri:trans}
%%     {\goesto[{\nopri[C]}]{}} \bydef{=}
%%     \bsetdef{\tuple{s, a(\true), g, u, s'}}{
%%       \tuple{s, a, g, u, s'} \in \goesto{}}
%%     \\    
%%     \cup
%%     \bsetdef{\tuple{s, a(\false), \true, \noop, s}}{
%%       s \in Q_C, a \in P_C}
%%     %% \\
%%     \cup
%%     \bigl\{(s_0, \init, \true, \mathit{init}, q^0_C)\bigr\}
%%     \,.
%%   \end{multline}

\item let  $\mathit{uinit} \bydef{=} \bigl(x := \val{0}{C}(x)\bigr)^{x \in V_C}$ and,
  %% \noteSB{Ludo, Eric, pourriez-vous v\'erifier que cela est
  %%   compr\'ehensible ? Ok pour moi (Eric) OK Ludo, rephrase un peu}
  %
  for all $s \goesto{a, g, u} s'$ with $u = (x\! :=
\! e_x)^{x
    \in V}$ (with $V \!\subseteq\! V_C$),
%  
  let $u' \bydef{=}
  \bigl(x := e_x\bigr)^{x \in V \setminus \export[C](a)}
  \cup
  \bigl(x := e_x\bigl[\nicefrac{?x'}{x}\bigr]\bigr)^{x \in V \cap \export[C](a)}$,
%
  and $\export[C]'(a) \bydef{=} \setdef{?x'}{x \in \export[C](p), p \in a}$,
%
%% \noteEM{j'ai inverse les $\export[C]'(a), \export[C](a)$ dans les
%%   actions, pour etre conforme a la notation de la Def 1 page 4. }
\begin{multline*}
    {\goesto[{\nopri[C]}]{}} \bydef{=} 
    \Setdef{\Tuple{s, a\bigl(\export[C]'(a), \export[C](a),\true\bigr), g, u', s'}}{
      s \goesto{a, g, u} s'}
    \\
    {} \cup
    \Setdef{\Tuple{s, a\bigl(\export[C]'(a), \export[C](a),\false\bigr), \true, \noop, s}}{
      s \in Q_C, \exists s' \in Q_C: s' \goesto{a}}
    \\
    {} \cup \bigl\{(s_0, \init, \true, \mathit{uinit}, q^0_C)\bigr\}.
  \end{multline*}
\end{itemize}%\vspace{-3ex}

@@

Basic research on behaviour models and verification algorithms for
data-sensitive systems started in the nineties, with the seminal work
of Hennesy, Lin, and their colleagues on value-passing systems with
assignments
\cite{CONCUR::Lin1996,HennessyL1995:353,APSEC::Lin2001}. Later, many
different works addressing various classes of infinite-state systems
and/or paramaterized topologies have been published, using
combinations of approaches, often including predicate abstraction and
SMT satisfiability (see
e.g. \cite{BruniEtAl-Tiles-Concur2000,DBLP:conf/cade/GhilardiNRZ08,DBLP:journals/jsat/AlbertiGPRR12,CimattiEtAl-NUXMV-CAV2014,ChampionEtAl-Kind2-CAV2016}). With
respect to these, we use symbolic representations not only to get a
finite representation of infinite spaces, but also to express the
(data-sensitive) synchronisations with the environment, making our
models suitable for compositional verification.
Among these works, several have shown the capacity of the SMT engines
(either Z3 or Yikes) as servers for solving verification conditions of
the algorithms, for large case-studies
(e.g. \cite{DBLP:journals/corr/CimattiGMT13,DBLP:journals/corr/abs-1806-11459}). %% Also

%% SMT on an application not using component structure. is there a strong Z3 example == Eric

@@

  Denoting %% \todoSB{One tilde is enough}
  $\val[\doubletilde]{}{1}$ the restriction of
%
  $\val[\tilde]{}{}$ to
  $V_1 = V_{C_1} \cup \bigcup_{B \in \cB} V_B$,
%
  we have, by \eq{substitution} and the monotonicity of
  expressions,
%
  $(\val[\primeit]{}{1}, \val[\primeit]{}{}) \order
  \val[\doubletilde]{}{1}[e^1; (e_1, e_b)]$.  %% \todoSB{Remove this}\hl{Notice that all
  %% variables affected by both $e^1$ and $e^2$ must necessarily
  %% belong to $\bigcup_{B \in \cB} V_B$.}
  Hence,
%
  $\val[\doubletilde]{}{1}[e^1; (e_1, e_b)] =
  (\val[\primeit]{}{1}, \val[\secondit]{}{})$ with 
  $\val[\primeit]{}{} \order \val[\secondit]{}{}$.

@@

  In particular, the first safety property \eq{prop:reset:can}:
  \begin{equation*}
    \AG \bigl(\PortFail(true) \rightarrow \EX \EU[]{\lnot \PortFail(true)}{\PortReset}\bigr).
  \end{equation*}
is not true, because the $OT_2$ loop on state $\ostate{11}$ stays in
state 11 where you need another $\PortFail(true)$ before enabling a
reset. This is because without priorities we can have an
activity on the dangling ``fail'' port without synchronising with the
others, even if they are available.

@@

%% \subsubsection{Scaling-up: composing architectures.}

%% Here is the code of the corresponding composed pNet:
%% %% \noteEMin{Il faudrait (mais a-t-on le temps?): faire un dessin de la
%% %%   composition au niveau BIP, et au niveau pNet. Un resultat brut, vaguement
%% %%   correct, montre 37 OTs, modulo les petites questions de mon mail de
%% %%   mercredi midi...}

%% %% \todoSBin{Ecrire les proprietes qu'il doit verifier... ? }

%% \bigskip
%% \begin{verbatim}
%% SV0: <tick(z, t, z1), fail(zone1, b11), fail(zone2, b12), fail1(b01), fail2(b02), _> 
%%      --> fail(b01, b02), 
%%      [(z1 = z |Intersection| ((b11?zone1:TopInt)+t) |Intersection| ((b12?zone2:TopInt)+t))
%%       /\(b11=>b01)/\(b12=>b02)]
%% SV1: <cancel(b1), resume(b21), resume(b22), resume1, resume2, _>-->resume, 
%%      [(b1=b21)/\(b1=b22)]
%% SV2: <timeout, timeout, timeout, _, _, ask>-->ask
%% SV3: <_, reset, reset, _, _, reset>-->reset
%% SV4: <init, init, init, _, _, _>-->init

%% pLTS: Timer
%% - Transition: t0---init-->t1, { t := 0 z := TopInt}
%% - Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
%% - Transition: t1---cancel(false)-->t1, {}
%% - Transition: t1---cancel(true)-->t1, { t := 0 z := TopInt}
%% - Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}

%% pLTS: FC1
%% - Transition: s0---init-->s1, { zone1 := minZone1,maxZone1}
%% - Transition: s1---timeout-->s1, {}
%% - Transition: s1---resume(false)-->s1, {}
%% - Transition: s1---fail(zone1, false)-->s1, {}
%% - Transition: s1---fail(zone1, true)-->s2, {}
%% - Transition: s2---fail(zone1, false)-->s2, {}
%% - Transition: s2---resume(false)-->s2, {}
%% - Transition: s2---resume(true)-->s1, {}
%% - Transition: s2---timeout-->s3, {}
%% - Transition: s3---fail(zone1, false)-->s3, {}
%% - Transition: s3---timeout-->s3, {}
%% - Transition: s3---resume(false)-->s3, {}
%% - Transition: s3---reset-->s1, {}

%% pLTS: FC2
%% - Transition: s0---init-->s1, { zone2 := minZone2,maxZone2}
%% - Transition: s1---timeout-->s1, {}
%% - Transition: s1---resume(false)-->s1, {}
%% - Transition: s1---fail(zone2, false)-->s1, {}
%% - Transition: s1---fail(zone2, true)-->s2, {}
%% - Transition: s2---fail(zone2, false)-->s2, {}
%% - Transition: s2---resume(false)-->s2, {}
%% - Transition: s2---resume(true)-->s1, {}
%% - Transition: s2---timeout-->s3, {}
%% - Transition: s3---fail(zone2, false)-->s3, {}
%% - Transition: s3---timeout-->s3, {}
%% - Transition: s3---resume(false)-->s3, {}
%% - Transition: s3---reset-->s1, {}

%% INFO  BIPArchFailureTimerMaxV4X2 - Total: 37 SATISFIABLE OTs
%% INFO  BIPArchFailureTimerMaxV4X2 -        3136 UNSATISFIABLE OTs
%% INFO  BIPArchFailureTimerMaxV4X2 - Total Time: 9111 ms
%% \end{verbatim}

%% \bigskip
%% \Fig{ArchFailure:OA-TimerV4X2} shows the open automata of
%% this composed architecture.
%% Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
%% ``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
%% false. As in the previous case, this is because we have not use here
%% the Max Progress hypothesis, otherwise all the ``*'' in the figure
%% should be replaced by ``$ff$''.

%%   \begin{figure}[t]
%%   \centering
%%   \includegraphics[width=0.8\columnwidth]{TimerV4X2}
%%   \caption{The open Automaton of the composed system}
%%   \label{fig:ArchFailure:OA-TimerV4X2}
%% \end{figure}
